---
title: "Introduction to the scFeatureFilter package"
author: "Vignette Authors"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to scFeatureFilter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Preamble
## Package aim
One early step of single cell RNA-seq analysis is the filtering of features (i.e. genes or transcripts)
to retain only the most expressed ones. This allows to focus on features less affected by technical
variations (relatively), and often increases the statistical power of the subsequent analyses.

This step is usually done by applying one or more arbitrary thresholds (i.e. mean TPM > 1, retention of features 
frequently detected across cells, etc.), or by using data from spike-in RNA.
This package aims to inform that threshold decision, especially in the cases
when spike-in controls are not available.

## Method summary
Highly expressed features are less affected by technical variability that lowly expressed ones,
and thus capture biological information more successfully.
We propose to use this property to set a small bin of highly expressed features as a reference to 
estimate the relative decrease in the biological variation component across the results. Namely, we
aim to evaluate the strength of the reference correlation to bins of features of decreasing expression.
In this scenario, it is expected that features from major transcription programs will be correlated 
with each other across bins, unless the technical variation component dominates the expression results.

One of the package's first steps is to select this reference bin of features (or 'top window' of features).
We propose to use the 100 features with the highest mean expression across cells, and a reasonably low
fraction of 0 expression as reference. Using more features will dilute the amount of biological variation
reflected in this reference bin (as well as increase computation time), while using less features might 
result in insufficient capture of the biological complexity of the samples.
Subsequent bins of features of decreasing mean expression across cells are then defined (by default, 1000 features
per bin). Note that the size of the subsequent feature bins matters much less than the size of the
reference bin.

For comparison, several negative control bins are defined by randomly shuffling values of the top bin expression matrix,
building a window that is expected to be completely uncorrelated. Then, every feature in each bin is correlated 
to every feature in the reference bin and the control bins to obtain correlation coefficient distributions.
Ultimately, the variation of the distributions of correlation coefficients can be used to inform the threshold decision.

## Installation
Instructions to arrive shortly.

Once installed, the package is ready to be loaded:
```{r}
library(scFeatureFilter)

library(ggplot2)
library(cowplot) # multipanel figures + nice theme
```


# Integrated usage
A single-function call option is available to the user, in which the entire package functionality is
executed. This will filter the expression matrix and keep only the most informative highly expressed features,
using default options:
```{r}
# example dataset included with the package:
scData_hESC

# filtering of the dataset with a single function call:
sc_feature_filter(scData_hESC)
```


# Detailed usage
## Loading data
*scFeatureFilter* uses an expression matrix (either as a `matrix`, a `data.frame`, a `tibble` or a
`SingleCellExperiment` R object) as input. Support of Bioconductor's expression set objects will arrive shortly.
Note that features should be in rows, while cells should be in columns. 


We recommend providing normalized expression values where at least library size is accounted for, 
such as TPM or FPKM, rather than raw counts.


An example dataset is supplied with the package:
```{r}
scData_hESC
```
*The example dataset was processed by [Mantsoki et al. (2016)](https://www.ncbi.nlm.nih.gov/pubmed/26951854),  data from human Embryonic Stem cells was generated by [Yan et al. (2013)](https://www.ncbi.nlm.nih.gov/pubmed/23934149).*

## Binning data
Firstly, a mean expression and a coefficient of variation column is added to the expression matrix provided
using the `calculate_cvs` function:
```{r}
calculate_cvs(scData_hESC)
```

We can explore the mean - variance relationship of the dataset with the `plot_mean_variance` function:
```{r collapse=TRUE}
library(magrittr) # to use the pipe %>%

calculate_cvs(scData_hESC) %>%
    plot_mean_variance(colourByBin = FALSE)
```

The top window (or reference bin) is then defined with the `define_top_genes` function, while the rest of bins
are created with the `bin_scdata` function:
```{r collapse=TRUE}
scData_hESC %>%
    calculate_cvs %>%
    define_top_genes(window_size = 100) %>%
    bin_scdata(window_size = 1000)
```

Then, we can plot the resulting bin definition using the same `plot_mean_variance` function, setting the 
`colourByBin` argument to `TRUE`:
```{r collapse=TRUE}
myPlot <- scData_hESC %>%
    calculate_cvs %>%
    define_top_genes(window_size = 100) %>%
    bin_scdata(window_size = 1000) %>%
    plot_mean_variance(colourByBin = TRUE, density_color = "blue")

myPlot
```

Note that the plotting function outputs are `ggplot2` objects that can be 
further customized. See an example here:
```{r}
myPlot + annotation_logticks(sides = "l")
```

## Correlation of the bins
After binning the data, correlations can be calculated. Every feature in each window is correlated 
to every other feature in the reference bin (as well as in the randomized control bins) using the 
function `correlate_windows`:
```{r collapse=TRUE}
corDistrib <- scData_hESC %>%
    calculate_cvs %>%
    define_top_genes(window_size = 100) %>%
    bin_scdata(window_size = 1000) %>%
    correlate_windows(n_random = 3)
```

For visualization of the correlation values, probability distributions can be computed using 
and plotted using the `correlations_to_densities` and `plot_correlations_distributions` functions:
```{r collapse=TRUE}
corDens <- correlations_to_densities(corDistrib, absolute_cc = TRUE)
plot_correlations_distributions(corDens, facet_ncol = 5)
```

The coloured lines indicate the correlation coefficient distributions for each bin compared to 
the reference bin (bin 1 shows the autocorrelation of the reference bin). The thicker blue 
lines with grey area (barely visible) represents the correlation coefficient distributions for
each bin compared to the randomized control bins.


Notice how the correlation distributions shift from high to low values as the bin number increases,
i.e. as the  mean expression level of the features decreases, indicating a higher influence of 
technical variability in the expression results.

*Note: the `absolute_cc = TRUE` option, which is the default, instructs the package to work
with the absolute value of the correlation coefficient. Therefore, highly negatively correlated features
will count as highly correlated. Setting this option as `TRUE` also produces clearer plots, and reduces 
the emphasis on non-symmetrical, near 0, shifts of correlation values that are not interpretable.*

We can quantify the steady decrease of correlation values as feature expression goes down using the
`get_mean_median` function, improving interpretability of the extent of technical variability in the data:
```{r}
metrics <- get_mean_median(corDistrib)
metrics
plot_correlations_distributions(corDens, metrics = metrics, facet_ncol = 5)
```

The added dashed lines are the mean values of the distributions. The `plot_metric` function focalized on these means of distributions:
```{r}
plot_metric(metrics, show_ctrl = FALSE, show_threshold = FALSE)
```

The bar represents the mean value of the correlation coefficient distribution of each bin 
against the reference bin, while the dot and whiskers are against the randomized control windows.
In our example dataset, the mean of the correlation distributions decreases steadily, 
and reaches a plateau after bin 11.

## Choosing a threshold
On the previous plot, a background reference can be added as a dashed line, by taking the mean value of the 
correlations of every bin to the randomized control bins:
```{r}
plot_metric(metrics, show_ctrl = TRUE, show_threshold = FALSE)
```

Then, the user can decide to plot only bins for which the mean correlation value against the reference window 
is higher than, for example, twice the background:
```{r}
plot_metric(metrics, show_ctrl = TRUE, show_threshold = TRUE, threshold = 2)
```

To further assist thresholding, the function `determine_bin_cutoff` will return the number of the first bin 
of features that falls below the set threshold:
```{r}
determine_bin_cutoff(metrics, threshold = 2)
```

Finally, using this information, the function `filter_expression_table` can be used to obtain a filtered 
expression matrix from the original data:
```{r}
binned_data <- scData_hESC %>%
    calculate_cvs %>%
    define_top_genes(window_size = 100) %>%
    bin_scdata(window_size = 1000)
metrics <- correlate_windows(bined_data, n_random = 3) %>%
    get_mean_median

filtered_data <- filter_expression_table(
    bined_data,
    bin_cutoff = determine_bin_cutoff(metrics)
)

dim(scData_hESC)
dim(filtered_data)
filtered_data

```

*Note: most of the features are filtered at the `calculate_cvs` step because their fraction of zero expression 
(i.e. the number of cells in which they present a 0 expression value) is prohitively high for CV calculation.
The maximum fraction of 0 expression tolerated by scFeatureFilter is 75%, by default.*

## Working with SingleCellExperiment objects

This package accpets `SingleCellExperiment` as inputs for `sc_feature_filter`
and `calculate_cvs` functions. However, it does not retun such objects at the moment.
One can retrieve a filtered `SingleCellExperiment` objects like this:
```{r}
library(SingleCellExperiment)
library(scRNAseq) # example datasets
data("allen")
sce_allen <- as(allen, "SingleCellExperiment")

# sce_allen is an SingleCellExperiment object
sce_allen

filtered_allen <- sc_feature_filter(sce_allen, sce_assay = "rsem_tpm")
is.matrix(filtered_allen) # filtered_allen is a tibble

sce_filtered_allen <- sce_allen[rownames(filtered_allen), ]
sce_filtered_allen
```

# Parameter values
Several parameters are used in `sc_feature_filter` or in step-by-step functions.
Their default values should be fine for eukaryote species with multi-exonic
genes (metazoans & plants).

## `max_zeros`
The `max_zeros` parameter default value `0.75`) is used as a first filter.
Features with more than this proportion of cells with 0 expression will be
removed. 0 expression can reflect both an absence of expression in the cell
of origin, or a drop-out event during the library preparation. This technical
drop out can be quite high in scRNAseq protocols. More stringent values for
this parameter could be `0.5` or `0.25`. As features with many 0-expression
events are usually lowly expressed in other cells, this parameter will 
mostly impact the number of bins in the rest of the method,
adding or removing lowly expressed bins of features.


## `top_window_size`
The `top_window_size` parameter (default to `100`) can have significant impacts
on the rest of the pipeline. It is the number of feature with the highest mean
expression that will be used as a reference bin. If the reference window is too
small (i.e. `10`), it will not capture enough of the biological variation.
On the contrary, a too large window (i.e. `2000`) might include to many features
highly affected from technical variations to be of any use. We therefore suggest
keeping it at `100` or near that value.

The function `plot_top_window_autocor` can be used to visualy explore the mean
auto-correlation value of features in the top bin with themselves depending on
the number of features in the top bin:
```{r}
plot_top_window_autocor(calculate_cvs(scData_hESC))
```


## `other_window_size`
The `other_window_size` parameter (default to `1000`) is the number of 
features included in the other bins of feature. It can safely be decreased to
`500` or `250`, which will result in more bins, without changing much the
proposed filtering threshold:
```{r}
metrics_bigBins <- scData_hESC %>%
    calculate_cvs %>%
    define_top_genes(window_size = 100) %>%
    bin_scdata(window_size = 1000) %>%
    correlate_windows(n_random = 3) %>%
    get_mean_median

metrics_smallBins <- scData_hESC %>%
    calculate_cvs %>%
    define_top_genes(window_size = 100) %>%
    bin_scdata(window_size = 500) %>%
    correlate_windows(n_random = 3) %>%
    get_mean_median

plot_grid(
    plot_metric(metrics_bigBins) +
        labs(title = "1000 genes per bin"),
    plot_metric(metrics_smallBins) +
        labs(title = "500 genes per bin")
)

```

## `n_random`
The background correlation values used to determine a threshold come from
shuffling of the values in the top bin expression matrices. `n_random`
(default to 3) is the control bins that will be generated. Result difference
between the different random windows are small, there is not much need to
increase that value. Increasing it a lot will increase computation time.

## `threshold`
One can decide which bin of features one want to keep looking at the
`plot_metric` or `plot_correlations_distributions` plots.
We also provide an automatic threshold decision, keeping only windows of
feature with a mean correlation value with the top window higher than 
`threshold` (default to 2) times the background level, estimated using 
randomised top bins. Depending on the downstream applications, one
can increase the `threshold` value to be more stringent in the selection.
